---
title: "RoB 2.0 automation"
output: html_notebook
---


# Housekeeping

```{r}
# install.packages(c("httr2", "jsonlite", "dplyr", "purrr", "readr", "usethis"))
library(httr2)
library(jsonlite)
library(dplyr)
library(purrr)
library(readr)
library(usethis)
```

# API configuration

```{r}
usethis::edit_r_environ() #loads API key to env var

api_key  <- Sys.getenv("OPENAI_API_KEY")
api_url  <- "https://api.openai.com/v1/chat/completions"
model_id <- "gpt-5.1"

if (api_key == "") stop("Please set OPENAI_API_KEY env var or hard-code api_key.")

```

# Extract text from PDF

Option to automatize this task from a .bib file or directory

```{r}
pdf_path <- "./test.pdf" #sets path
output_path <- "./test.txt"
text_all <- pdf_text(pdf_path) #function pdf_text does the task
text_pages <- pdf_text(pdf_path) #returns character vector, one element per page
writeLines(text_pages, output_path
          )

# cat(substr(text_all, 1, 500)) #prints first 500 characters
```

# Load text

```{r}
trial_text <- read_file("./test.txt")
```

# 1. Call LLM function

Calls a LLM through HTTP

```{r}
call_llm <- function(system_prompt, user_prompt,
                     model = model_id,
                     temperature = 0) {
  req <- request(api_url) |>
    req_auth_bearer_token(api_key) |>
    req_body_json(list(
      model = model,
      messages = list(
        list(role = "system", content = system_prompt),
        list(role = "user",   content = user_prompt)
      ),
      temperature = temperature
    ))
  
  resp <- req_perform(req)
  cont <- resp_body_json(resp)
  
  # Basic safety checks
  if (!is.null(cont$error)) {
    stop(paste("LLM API error:", cont$error$message))
  }
  
  out <- cont$choices[[1]]$message$content
  out
}
```

## Testing LLM call

```{r}
result <- call_llm(
  system_prompt="you are a concise assistant.",
  user_prompt="Explain the role of dopamine in coding in 3 sentences."
  )
cat(result)
```

# 2. Master Prompt

Initial prompt needs to be created and adjusted according to the RoB version.
How to categorize RCT so version can be chosen automatically?
How to load RoB 2.0 into LLM's context?

```{r}
rob2_system_prompt <- "
You are assisting with bias assessment using the Risk of Bias 2.0 (RoB 2.0) tool
for cluster-randomized, parallel-group trials.

Rules:
1. Use only these response categories for signalling questions:
   Y (Yes), PY (Probably Yes), PN (Probably No), N (No), NI (No Information), NA (Not Applicable, only if allowed).
2. Apply the RoB 2.0 domains for cluster RCTs:
   - Domain 1a: Bias arising from the randomisation process.
   - Domain 1b: Bias from timing of identification/recruitment of participants.
   - Domain 2: Bias due to deviations from intended interventions.
   - Domain 3: Bias due to missing outcome data.
   - Domain 4: Bias in measurement of the outcome.
   - Domain 5: Bias in selection of the reported result.
3. For each domain:
   - Use only the information provided.
   - Answer every signalling question.
   - Provide a domain judgement: 'Low risk', 'Some concerns', or 'High risk'.
   - Provide a short, factual justification (2–4 sentences).
   - If the trial does not report something, use NI.
4. Do not invent information. If unsure, choose NI.
5. Overall RoB (computed by the user) will follow the usual RoB 2 algorithm; focus on domain-level outputs and justifications.
Output MUST be valid JSON whenever JSON is requested.
"
```

# 3. Extraction prompt: compress trial text into structured summary

```{r}
build_extraction_prompt <- function(trial_text) {
  sprintf("
Extract from the trial report only the information relevant to risk-of-bias
assessment in cluster-randomised, parallel-group trials.

Using only the information in the text, produce JSON with this structure:

{
  \"randomisation_and_allocation\": \"...\",
  \"cluster_recruitment_timing\": \"...\",
  \"blinding_and_deviations\": \"...\",
  \"attrition_and_missing_data\": \"...\",
  \"outcome_measurement\": \"...\",
  \"selective_reporting_indicators\": \"...\"
}

Each field should be a concise paragraph (<= 200 words), paraphrasing or quoting
relevant parts. If there is no information for a field, write \"No information reported\".

Trial text:
\"\"\"%s\"\"\"",
          trial_text
  )
}
```

# 4. Domain prompts

## Domain 1a

```{r}
build_domain1a_prompt <- function(text_block) {
  sprintf("
Using the RoB 2.0 cluster-randomised rules, evaluate Domain 1a:
Bias arising from the randomisation process.

Use ONLY the information in the text below.

Return JSON:

{
  \"domain\": \"1a_randomisation_process\",
  \"signalling_questions\": {
    \"1a.1\": \"\",
    \"1a.2\": \"\",
    \"1a.3\": \"\"
  },
  \"judgement\": \"\",
  \"rationale\": \"\"
}

Where:
- 1a.1: Was the allocation sequence random?
- 1a.2: Was the allocation sequence concealed until clusters were enrolled and assigned?
- 1a.3: Did baseline cluster/participant characteristics suggest successful randomisation?

Use only: Y, PY, PN, N, NI.

Text:
\"\"\"%s\"\"\"",
          text_block
  )
}
```

## Domain 1b

```{r}
build_domain1b_prompt <- function(text_block) {
  sprintf("Evaluate Domain 1b: Bias from timing of identification and recruitment of participants
in a cluster-randomised trial.

Use ONLY the information in the text below.

Return JSON:

{
  \"domain\": \"1b_timing_of_identification_recruitment\",
  \"signalling_questions\": {
    \"1b.1\": \"\",
    \"1b.2\": \"\",
    \"1b.3\": \"\"
  },
  \"judgement\": \"\",
  \"rationale\": \"\"
}

Where:
- 1b.1: Were all participants identified before randomisation of clusters?
- 1b.2: If not, could knowledge of cluster assignment influence participant recruitment?
- 1b.3: Is there baseline imbalance consistent with biased recruitment?

Use Y, PY, PN, N, NI.

Text:
\"\"\"%s\"\"\"",
          text_block
  )
}
```

## Domain 2

```{r}
build_domain2_prompt <- function(text_block) {
  sprintf("
Evaluate Domain 2: Bias due to deviations from intended interventions.

Use ONLY the information in the text below.

Return JSON:

{
  \"domain\": \"2_deviations_from_intended_interventions\",
  \"signalling_questions\": {
    \"2.1\": \"\",
    \"2.2\": \"\",
    \"2.3\": \"\",
    \"2.4\": \"\",
    \"2.5\": \"\"
  },
  \"judgement\": \"\",
  \"rationale\": \"\"
}

Where (generic RoB 2 wording):
- 2.1: Were participants aware of their assigned intervention?
- 2.2: Were caregivers or personnel aware?
- 2.3: Were there deviations from intended intervention that arose because of the experimental context?
- 2.4: Were these deviations likely to have affected the outcome, and were they unbalanced between groups?
- 2.5: Was the analysis appropriate to estimate the effect of assignment (e.g. intention-to-treat)?

Use Y, PY, PN, N, NI.

Text:
\"\"\"%s\"\"\"",
          text_block
  )
}
```

## Domain 3

```{r}
build_domain3_prompt <- function(text_block) {
  sprintf("
Evaluate Domain 3: Bias due to missing outcome data.

Use ONLY the information in the text below.

Return JSON:

{
  \"domain\": \"3_missing_outcome_data\",
  \"signalling_questions\": {
    \"3.1\": \"\",
    \"3.2\": \"\",
    \"3.3\": \"\"
  },
  \"judgement\": \"\",
  \"rationale\": \"\"
}

Where:
- 3.1: Is the proportion of missing outcome data low?
- 3.2: Is missingness of the outcome data unlikely to depend on its true value?
- 3.3: Were appropriate methods used to handle missing data?

Use Y, PY, PN, N, NI.

Text:
\"\"\"%s\"\"\"",
          text_block
  )
}
```


## Domain 4

```{r}
build_domain4_prompt <- function(text_block) {
  sprintf("
Evaluate Domain 4: Bias in measurement of the outcome.

Use ONLY the information in the text below.

Return JSON:

{
  \"domain\": \"4_measurement_of_outcome\",
  \"signalling_questions\": {
    \"4.1\": \"\",
    \"4.2\": \"\",
    \"4.3\": \"\",
    \"4.4\": \"\"
  },
  \"judgement\": \"\",
  \"rationale\": \"\"
}

Where:
- 4.1: Was the method of measuring the outcome appropriate?
- 4.2: Could measurement or ascertainment of the outcome have differed between intervention groups?
- 4.3: Were outcome assessors aware of the intervention received by participants?
- 4.4: Could assessment of the outcome have been influenced by knowledge of intervention received?

Use Y, PY, PN, N, NI.

Text:
\"\"\"%s\"\"\"",
          text_block
  )
}
```

## Domain 5

```{r}
build_domain5_prompt <- function(text_block) {
  sprintf("
Evaluate Domain 5: Bias in selection of the reported result.

Use ONLY the information in the text below, which may include trial registry or protocol information if provided.

Return JSON:

{
  \"domain\": \"5_selection_of_reported_result\",
  \"signalling_questions\": {
    \"5.1\": \"\",
    \"5.2\": \"\",
    \"5.3\": \"\"
  },
  \"judgement\": \"\",
  \"rationale\": \"\"
}

Where:
- 5.1: Is the reported analysis for the outcome consistent with a prespecified analysis plan?
- 5.2: Are all prespecified outcomes for this domain reported?
- 5.3: Is there evidence of selective reporting of analyses or outcomes?

Use Y, PY, PN, N, NI.

Text:
\"\"\"%s\"\"\"",
          text_block
  )
}
```

# 5. Overall Risk of Bias

```{r}
overall_rob_from_domains <- function(domain_judgements) {
  # domain_judgements: character vector (e.g. c("Low risk", "Some concerns", "High risk", ...))
  dom <- tolower(domain_judgements)
  
  if (any(dom == "high risk")) {
    overall <- "High"
  } else if (all(dom == "low risk")) {
    overall <- "Low"
  } else {
    overall <- "Some concerns"
  }
  
  overall
}
```

# 6. SERVER

### Function to assess one study

```{r}
assess_rob2_for_trial <- function(study_id,
                                  outcome,
                                  trial_text,
                                  verbose = TRUE) {
  if (verbose) message("Assessing study: ", study_id, " | outcome: ", outcome)
  extraction_prompt <- build_extraction_prompt(trial_text)
  extraction_raw    <- call_llm(rob2_system_prompt, extraction_prompt)
  
  extraction_json <- tryCatch(
    fromJSON(extraction_raw),
    error = function(e) {
      warning("Failed to parse extraction JSON for study ", study_id, ": ", e$message)
      list(
        randomisation_and_allocation   = trial_text,
        cluster_recruitment_timing     = trial_text,
        blinding_and_deviations        = trial_text,
        attrition_and_missing_data     = trial_text,
        outcome_measurement            = trial_text,
        selective_reporting_indicators = trial_text
      )
    }
  )

  dom1a_prompt <- build_domain1a_prompt(extraction_json$randomisation_and_allocation)
  dom1a_raw    <- call_llm(rob2_system_prompt, dom1a_prompt)
  dom1a        <- fromJSON(dom1a_raw)
  
  dom1b_prompt <- build_domain1b_prompt(extraction_json$cluster_recruitment_timing)
  dom1b_raw    <- call_llm(rob2_system_prompt, dom1b_prompt)
  dom1b        <- fromJSON(dom1b_raw)
  
  dom2_prompt  <- build_domain2_prompt(extraction_json$blinding_and_deviations)
  dom2_raw     <- call_llm(rob2_system_prompt, dom2_prompt)
  dom2         <- fromJSON(dom2_raw)
  
  dom3_prompt  <- build_domain3_prompt(extraction_json$attrition_and_missing_data)
  dom3_raw     <- call_llm(rob2_system_prompt, dom3_prompt)
  dom3         <- fromJSON(dom3_raw)
  
  dom4_prompt  <- build_domain4_prompt(extraction_json$outcome_measurement)
  dom4_raw     <- call_llm(rob2_system_prompt, dom4_prompt)
  dom4         <- fromJSON(dom4_raw)
  
  dom5_prompt  <- build_domain5_prompt(extraction_json$selective_reporting_indicators)
  dom5_raw     <- call_llm(rob2_system_prompt, dom5_prompt)
  dom5         <- fromJSON(dom5_raw)
  
 
  domain_judgements <- c(
    dom1a$judgement,
    dom1b$judgement,
    dom2$judgement,
    dom3$judgement,
    dom4$judgement,
    dom5$judgement
  )
  
  overall <- overall_rob_from_domains(domain_judgements)
  
  list(
    study_id   = study_id,
    outcome    = outcome,
    extraction = extraction_json,
    domains    = list(
      d1a = dom1a,
      d1b = dom1b,
      d2  = dom2,
      d3  = dom3,
      d4  = dom4,
      d5  = dom5
    ),
    overall_rob = overall
  )
}
```



```{r}
extraction_prompt <- build_extraction_prompt(trial_text)
extraction_raw    <- call_llm(rob2_system_prompt, extraction_prompt)
```
Concatenates extraction_prompt and trial_text
Calls LLM, sends main prompt and extraction prompt
extraction_raw = LLM's answer

```{r}
extraction_json <- tryCatch(
    fromJSON(extraction_raw),
    error = function(e) {
      warning("Failed to parse extraction JSON for study ", study_id, ": ", e$message)
      list(
        randomisation_and_allocation   = trial_text,
        cluster_recruitment_timing     = trial_text,
        blinding_and_deviations        = trial_text,
        attrition_and_missing_data     = trial_text,
        outcome_measurement            = trial_text,
        selective_reporting_indicators = trial_text
      )
    }
)
```
Parses info into a JSON from extraction, else pastes whole trial_text in each field

```{r}
dom1a_prompt <- build_domain1a_prompt(extraction_json$randomisation_and_allocation)
dom1a_raw <- call_llm(rob2_system_prompt, dom1a_prompt)
dom1a <- fromJSON(dom1a_raw)
```
Uses pertinent text (summarized by LLM) to send a new LLM call with domain-specific prompt, then converts output to JSON

```{r}
dom1b_prompt <- build_domain1b_prompt(extraction_json$cluster_recruitment_timing)
dom1b_raw    <- call_llm(rob2_system_prompt, dom1b_prompt)
dom1b        <- fromJSON(dom1b_raw)
  
dom2_prompt  <- build_domain2_prompt(extraction_json$blinding_and_deviations)
dom2_raw     <- call_llm(rob2_system_prompt, dom2_prompt)
dom2         <- fromJSON(dom2_raw)
  
dom3_prompt  <- build_domain3_prompt(extraction_json$attrition_and_missing_data)
dom3_raw     <- call_llm(rob2_system_prompt, dom3_prompt)
dom3         <- fromJSON(dom3_raw)
  
dom4_prompt  <- build_domain4_prompt(extraction_json$outcome_measurement)
dom4_raw     <- call_llm(rob2_system_prompt, dom4_prompt)
dom4         <- fromJSON(dom4_raw)
  
dom5_prompt  <- build_domain5_prompt(extraction_json$selective_reporting_indicators)
dom5_raw     <- call_llm(rob2_system_prompt, dom5_prompt)
dom5         <- fromJSON(dom5_raw)
```
Same for each domain

```{r}
domain_judgements <- c(
    dom1a$judgement,
    dom1b$judgement,
    dom2$judgement,
    dom3$judgement,
    dom4$judgement,
    dom5$judgement
  )
```
Creates a vector with all domains

```{r}
overall <- overall_rob_from_domains(domain_judgements)
```
Applies RoB algorithm to vector

```{r}
list(
    study_id   = study_id,
    outcome    = outcome,
    extraction = extraction_json,
    domains    = list(
      d1a = dom1a,
      d1b = dom1b,
      d2  = dom2,
      d3  = dom3,
      d4  = dom4,
      d5  = dom5
    ),
    overall_rob = overall
  )
```
Creates a dataframe with summarized info, results for each domain, and overall judgement

# 7. Apply to a dataset of trials

```{r}

#trials_df <- read_csv("trials_texts.csv")

trials_df <- tibble::tibble(
  study_id  = c("ExampleTrial1"),
  outcome   = c("Primary outcome"),
  trial_text = c(
    "Example RCT text goes here."
  )
)
```
Loads dataset from CSV into a tibble trials_df

```{r}
rob_results <- pmap(
  trials_df,
  ~ assess_rob2_for_trial(
    study_id  = ..1,
    outcome   = ..2,
    trial_text = ..3,
    verbose   = TRUE
  )
)
```
pmap loops through each row

```{ruby}
#rob_results

List of 1
 $ :List of 5
  ..$ study_id   : chr "ExampleTrial1"
  ..$ outcome    : chr "Primary outcome"
  ..$ extraction :List of 6
  .. ..$ randomisation_and_allocation   : chr "The study used a computer-generated sequence..."
  .. ..$ cluster_recruitment_timing     : chr "Participants were recruited after cluster assignment..."
  .. ..$ blinding_and_deviations        : chr "The trial was open-label..."
  .. ..$ attrition_and_missing_data     : chr "10% attrition, balanced between groups..."
  .. ..$ outcome_measurement            : chr "Outcome assessors were blinded..."
  .. ..$ selective_reporting_indicators : chr "Protocol available in registry; all outcomes reported..."
  ..$ domains   :List of 6
  .. ..$ d1a:List of 4
  .. .. ..$ domain             : chr "1a_randomisation_process"
  .. .. ..$ signalling_questions:List of 3
  .. .. .. ..$ 1a.1 = Y
  .. .. .. ..$ 1a.2 = Y
  .. .. .. ..$ 1a.3 = Y
  .. .. ..$ judgement          : chr "Low risk"
  .. .. ..$ rationale          : chr "Randomisation was computer-generated and adequately concealed."
  .. ..$ d1b:List of 4
  .. ..$ d2 :List of 4
  .. ..$ d3 :List of 4
  .. ..$ d4 :List of 4
  .. ..$ d5 :List of 4
  ..$ overall_rob: chr "Some concerns"

```
rob_results is a *list* where each element is the full output of main server function

```{r}
rob_tbl <- tibble::tibble(
  study_id   = map_chr(rob_results, "study_id"),
  outcome    = map_chr(rob_results, "outcome"),
  overall_rob = map_chr(rob_results, "overall_rob"),
  extraction  = map(rob_results, "extraction"),
  domains     = map(rob_results, "domains")
)
```
creates a tibble from rob_results with one row per study+outcomes

```{r}
flatten_domains <- function(x) {
  c(
    d1a = x$d1a$judgement,
    d1b = x$d1b$judgement,
    d2  = x$d2$judgement,
    d3  = x$d3$judgement,
    d4  = x$d4$judgement,
    d5  = x$d5$judgement
  )
}

domain_matrix <- rob_tbl$domains |>
  map(flatten_domains) |>
  bind_rows()

rob_flat <- bind_cols(
  rob_tbl |> select(study_id, outcome, overall_rob),
  domain_matrix
)

print(rob_flat)
```
flattens the tibble

```{r}
# write_csv(rob_flat, "rob2_trials_results.csv")
```
Export to CSV